---
title: "HW2-exploration"
author: Madi Calbert
date: February 2, 2025
message: false
warning: false 
format:
  html:
    embed-resources: true 
---

# Set Up

```{r}
#| message: false
#| warning: false 

#load libraries
library(tidyverse)
library(here)
library(janitor)
library(scales)

#load data
cites <- read.csv(here("data/cites_felidae_1975_2024.csv"))
cites_recent <- read.csv(here("data/cites_felidae_2000_2024.csv"))
```

# Tidy Data

```{r}
#clean and wrangle data
cites <- cites %>% 
  clean_names() 

#add column of common names of top 10 species 
common_names <- data.frame(
  taxon = c("Panthera leo", 
            "Panthera tigris", 
            "Acinonyx jubatus", 
            "Panthera pardus", 
            "Lynx rufus", 
            "Puma concolor", 
            "Lynx lynx", 
            "Leptailurus serval", 
            "Caracal caracal", 
            "Lynx canadensis"),  
  common_name = c("Lion", 
                  "Tiger", 
                  "Cheetah", 
                  "Leopard", 
                  "Bobcat", 
                  "Mountain Lion", 
                  "Eurasian Lynx", 
                  "Serval", 
                  "Caracal", 
                  "Canada Lynx"))

cites <- cites %>%
  left_join(common_names, by = "taxon")  

missing_species <- cites %>%
  filter(is.na(common_name)) %>%
  distinct(taxon)

# print(missing_species)  #shows scientific names without a match, there's a lot lol
```


# Data Exploration

## Top 10 Reported Species

What is the most commonly reported species from the *Felidae* Family
in the CITES database?

```{r}
#data viz of most commonly reported species

#Count instances per Taxon
species_count <- cites %>%
  count(taxon)  

#Merge counts and common names
top_species <- species_count %>%
  left_join(common_names, by = "taxon")  

#Arrange and select top 10 species
top_species <- top_species %>%
  arrange(n) %>%                          
  slice_max(n, n = 10)                   

#Plot top 10 species
ggplot(top_species, aes(x = reorder(common_name, n), y = n)) +
  geom_col() +
  geom_text(aes(label = comma(n)), hjust = 1.2, color = "white") +
  coord_flip() +
  theme_minimal() +
  labs(y = "Individuals Reported", 
       x = NULL,
       title = expression("Top 10 CITES Reported Species in " * italic("Felidae") * " Family from 1975-2024")) +
   scale_y_continuous(expand = expansion(mult = c(0,0)), labels = comma)
```

## Importers / Exporters

Which countries are the largest importers/exporters for the top 10
species within the *Felidae* Family?

```{r}
#create new df to isolate the top 10 reported species from above
top_cats_2024 <- cites %>%
  filter(common_name %in% top_species$common_name) %>% 
  filter(year == 2024)

top_cats <- cites %>%
  filter(common_name %in% top_species$common_name)

#I want to try to make a map that shows the flows of exporting countries to importing countries
```


```{r}
# amount of live lions in the database, just was curious
cites %>% 
  filter(common_name == "Lion") %>% 
  filter(term == "live") %>% 
  count(taxon)  
```

## Purpose / Source

What is the most common purpose (captive breeding, hunting, circus,
etc.) for top 10 cat transactions?

The reported purpose of the transaction is shown as a one-letter code:

-   B Breeding in captivity or artificial propagation

-   E Educational

-   G Botanical garden

-   H Hunting trophy

-   L Law enforcement/judicial/forensic

-   M Medical (including biomedical research)

-   N Reintroduction or introduction into the wild

-   P Personal

-   Q Circus or travelling exhibition

-   S Scientific

-   T Commercial

-   Z Zoo

```{r}
top_cats %>%
  filter(year >= 2022 & year <= 2024) %>% 
  group_by(purpose) %>%
  ggplot(aes(x = common_name, fill = purpose)) +
  geom_bar(stat = "count") +
  # scale_fill_manual(
  #   labels = c("Breeding in captivity or artificial propagation",
  #              "Educational", 
  #              "Botanical garden", 
  #              "Hunting trophy",
  #              "Law enforcement/judicial/forensic", 
  #              "Medical (including biomedical research)",
  #              "Reintroduction or introduction into the wild",
  #              "Personal",
  #              "Circus or travelling exhibition",
  #              "Scientific",
  #              "Commercial",
  #              "Zoo")) +
  coord_flip() +
  theme(legend.position = "bottom")
```

Source codes:

-   C Animals bred in captivity in accordance with Resolution Conf.
    10.16 (Rev.), as well as parts and derivatives thereof, exported
    under the provisions of Article VII, paragraph 5

-   D Appendix-I animals bred in captivity for commercial purposes in
    operations included in the Secretariat’s Register, in accordance
    with Resolution Conf. 12.10 (Rev. CoP15), and Appendix-I plants
    artificially propagated for commercial purposes, as well as parts
    and derivatives thereof, exported under the provisions of Article
    VII, paragraph 4, of the Convention

-   F Animals born in captivity (F1 or subsequent generations) that do
    not fulfil the definition of ‘bred in captivity’ in Resolution Conf.
    10.16 (Rev.), as well as parts and derivatives thereof

-   I Confiscated or seized specimens (may be used with another code)

-   O Pre-Convention specimens R Ranched specimens: specimens of animals
    reared in a controlled environment, taken as eggs or juveniles from
    the wild, where they would otherwise have had a very low probability
    of surviving to adulthood

-   U Source unknown (must be justified)

-   W Specimens taken from the wild

```{r}
top_cats %>% 
  group_by(common_name, source) %>%
  ggplot(aes(x = common_name, fill = source)) +
  geom_bar() +
  coord_flip()
```

## Across Time

How has the number of wildlife traded within the Felidae family changed
over the years?

```{r}
felidae_trade_trend <- top_cats %>%
  group_by(year) %>%
  summarize(total_trade = n(), .groups = "drop")  # Count occurrences per year
```

```{r}
ggplot(felidae_trade_trend, aes(x = year, y = total_trade)) +
  geom_line(color = "blue", size = 1) +  
  geom_point(color = "red", size = 2) +  
  labs(title = "Wildlife Trade Trends in the Felidae Family (1975-2024)",
       x = "Year",
       y = "Total Transactions") +
  theme_minimal() 
```

2013: CITES CoP16 added new trade restrictions on big cats, pangolins,
and tropical birds.

2016: CITES CoP17 introduced stronger protections for species like
African lions (Panthera leo), with new regulations limiting the export
of wild lion parts (e.g., bones used in traditional medicine).

2019: CITES CoP18 imposed stricter trade regulations on cheetahs
(Acinonyx jubatus), frequently trafficked as exotic pets.

```{r}
felidae_trade_purpose <- top_cats %>%
  filter(year >= 2000) %>%  
  group_by(year, purpose) %>%
  summarize(total_trade = n(), .groups = "drop")

ggplot(felidae_trade_purpose, aes(x = year, y = total_trade, fill = purpose)) +
  geom_area(alpha = 0.8) +
  labs(title = "Purpose of Felidae Trade Over Time (2000 - Present)",
       x = "Year",
       y = "Total Transactions") +
  theme_minimal()
```

# Questions

1.  What have you learned about your data? Have any potentially
    interesting patterns emerged?

-    The top 10 species in the Felidae family revealed that lions were
    the most commonly reported species. My favorite animal is the
    cheetah and I initially wanted to focus on them but have realized
    that it maybe more interesting to consider the top 10 species.

-   By analyzing importers/exporters, I identified that the largest
    exporters are South Africa and Mozambique. Meanwhile the largest
    importer is Panama and US.

-   The purposes behind Felidae transactions, such as commercial use,
    captive breeding, or scientific purposes, might show a pattern of
    increasing demand for exotic pets (e.g., cheetahs as pets),
    especially after 2010. I want to dig into this more.

-   Looking at trade data from 1975-2024, revealed major shifts in the
    number of reported species. I identifying some of these trends to
    better assess how the wildlife trade has evolved.

2.  In HW #1, you outlined some questions that you wanted to answer
    using these data. Have you made any strides towards answering those
    questions? If yes, how so? If no, what next steps do you need to
    take (e.g. I need to create X plot type, I still need to track down
    Y data, I need to restructure existing data so that you can
    visualize it in Z ways, etc.)?

    -   I am pretty happy with my bar plot of the top 10 most commonly
        reported species, I could definitively make it look nicer
        thematically.

    -   I would like to make some sort of map to showcase the
        exporter/importer data.

    -   I would like to improve my time series plot to include more
        information about critical events or policy changes that
        impacted wildlife trade.

3.  What challenges do you foresee encountering with your data? These
    can be data wrangling and / or visualization challenges.

    -   I think that making the map of exporters to importers will be
        challenging. But I know I can do it with more time!

    -   I also was trying to figure out a better way to add a column of
        the common names and tried GBIF and other avenues but was
        unsuccessful.
